## README      

where "### !Manual assessment required" is flagged, pause the analysis and follow guidelines for e.g. reading literature.       
   
## Set Up and VEP  

PLINK binary bed, fam, bim files downloaded from 200,000 UK Biobank exome dataset and variants in the region (+/- 100bp) of genes of interest extracted into bfiles and a VCF file [ukbexomehcm.*]. For the same data, the frequency (--freq) and missingness information (--missing) were obtained via PLINK for the variants. A "metadata table" was created of ID's from the fam file and summary information (ukbbMetaData.txt).     

Run VCF by VEP (version 99) for the following parameters:   
```{bash VEP}
--symbol \
--canonical \
--mane \
--hgvs \
--custom clinvar_20201026.vcf.gz,ClinVar,vcf,exact,0,CLNSIG,CLNREVSTAT,CLNDN \
--custom gnomAD/r2.1.1_liftover38/exomes/gnomad.exomes.r2.1.1.sites.liftover_grch38.vcf.gz,gnomADg,vcf,exact,0,AF_afr,AF_amr,AF_asj,AF_eas,AF_fin,AF_nfe,AF_sas,AF_oth,AC,AC_popmax,AN,AN_popmax,AF,AF_popmax,popmax,faf95,faf95_nfe,faf95_afr,faf95_amr,faf95_eas,faf95_sas \
--plugin dbscSNV,dbscSNV1.1_GRCh38.txt.gz \
--plugin MaxEntScan \
--dir_plugins loftee/ \
--plugin LoF,\
loftee_path:loftee,\
human_ancestor_fa:human_ancestor.fa.gz,\
conservation_file:loftee.sql,\
gerp_bigwig:gerp_conservation_scores.homo_sapiens.GRCh38.bw \
--output_file ukb200exomehcmvepresults
```

## Packages    

```{r library}
library('data.table')
library('plyr')
library('tidyr')
library('dplyr')
library('expss')
library('ggpubr')
```

## Data Load

```{r files to load in}
vars <- fread("ukb200exomehcmvepresults", header=T, sep="\t")
vars <- data.frame(vars)

#count of missing (unmeasured) per variant in 50k
miss <- fread("ukb200exomehcmvepresults.lmiss", header=T)
miss <- data.frame(miss)
colnames(miss)[2] <- ("ID")

#minor allele frequency of the variants
freq <- fread("ukb200exomehcmvepresults.frq", header=T)
freq <- data.frame(freq)
colnames(freq)[2] <- ("ID")

bim <- fread("ukb200exomehcmvepresults.bim", header=F)
bim <- data.frame(bim)
colnames(bim) <- c("CHROM", "ID", "X", "POS", "ALT", "REF")

meta <- read.table("ukbbMetaData.txt", header=T)

fam <- fread("ukb200exomehcm.fam", header=F)
fam <- data.frame(fam)
```

## Identify protein altering variants   

Variants with a consequence term defined by the Sequence Ontology with the potential to alter the gene product level or structure:  

**transcript_ablation**	A feature ablation whereby the deleted region includes a transcript feature	HIGH	  
**splice_acceptor_variant**	A splice variant that changes the 2 base region at the 3' end of an intron	HIGH	  
**splice_donor_variant**	A splice variant that changes the 2 base region at the 5' end of an intron	HIGH	  
**stop_gained**	A sequence variant whereby at least one base of a codon is changed, resulting in a premature stop codon, leading to a shortened transcript	HIGH	  
**frameshift_variant**	A sequence variant which causes a disruption of the translational reading frame, because the number of nucleotides inserted or deleted is not a multiple of three	HIGH	  
**stop_lost**	A sequence variant where at least one base of the terminator codon (stop) is changed, resulting in an elongated transcript	HIGH	  
**start_lost**	A codon variant that changes at least one base of the canonical start codon	HIGH	  
**transcript_amplification**	A feature amplification of a region containing a transcript	HIGH	  
**inframe_insertion**	An inframe non synonymous variant that inserts bases into in the coding sequence	MODERATE	  
**inframe_deletion**	An inframe non synonymous variant that deletes bases from the coding sequence	MODERATE	  
**missense_variant**	A sequence variant, that changes one or more bases, resulting in a different amino acid sequence but where the length is preserved	MODERATE	  
**protein_altering_variant**	A sequence_variant which is predicted to change the protein encoded in the coding sequence	MODERATE	  
**regulatory_region_ablation**	A feature ablation whereby the deleted region includes a regulatory region	MODERATE	  

Plus variants with a consequence term requiring further assessment to subset those with the potential to alter the gene product level or structure:

**splice_region_variant**	A sequence variant in which a change has occurred within the region of the splice site, either within 1-3 bases of the exon or 3-8 bases of the intron	LOW	    
**incomplete_terminal_codon_variant**	A sequence variant where at least one base of the final codon of an incompletely annotated transcript is changed	LOW	    
**coding_sequence_variant**	A sequence variant that changes the coding sequence	MODIFIER	    
**start and stop codons** Require additional manual assessment       

```{r filter protein altering}
PAV_List=c("missense_variant","frameshift_variant","stop_gained","inframe_deletion","splice_donor_variant","splice_acceptor_variant","inframe_insertion","transcript_ablation","stop_lost","start_lost","transcript_amplification","protein_altering_variant", "splice_region_variant", "regulatory_region_ablation", "incomplete_terminal_codon_variant", "coding_sequence_variant")

ttn2 <- vars[grepl(paste(PAV_List, collapse="|"), vars$Consequence),]
```

## Filter for Genes of interest     

Adapted from Ingles et al., Circ Genom Precis Med. 2019;12:e002460 - ClinGen consortium     
8 definitive sarcomeric HCM genes = MYH7	MYBPC3	MYL2	MYL3	ACTC1	TNNI3	TNNT2	TPM1       
3 medium evidence HCM genes = CSRP3 TNNC1 JPH2    
2 intrinsic cardiomyopathy genes = ACTN2 (moderate) PLN (definitive)      
14 syndromic genes that can cause isolated LVH = FHL1 TTR	FLNC	GLA	LAMP2	PRKAG2	PTPN11	RAF1	RIT1	ALPK3	MYH7*	CACNA1C	DES     

```{r filter genes}
ttn2 <- ttn2[
ttn2$SYMBOL == "CSRP3" |
ttn2$SYMBOL == "PLN"  |
ttn2$SYMBOL == "TNNC1" |
ttn2$SYMBOL == "FHL1" |
ttn2$SYMBOL == "ACTN2"  |
ttn2$SYMBOL == "JPH2"  |
ttn2$SYMBOL == "FLNC"  |
ttn2$SYMBOL == "GLA"  |
ttn2$SYMBOL == "LAMP2"  |
ttn2$SYMBOL == "PRKAG2"  |
ttn2$SYMBOL == "TTR"  |
ttn2$SYMBOL == "PTPN11"  |
ttn2$SYMBOL == "RAF1" |
ttn2$SYMBOL == "RIT1"  |
ttn2$SYMBOL == "ALPK3" |
ttn2$SYMBOL == "CACNA1C"  |
ttn2$SYMBOL == "MYH7"  |
ttn2$SYMBOL == "DES"  |
ttn2$SYMBOL == "MYBPC3"  |
ttn2$SYMBOL == "TNNT2"  |
ttn2$SYMBOL == "TNNI3"  |
ttn2$SYMBOL == "TPM1"  |
ttn2$SYMBOL == "MYL2"  |
ttn2$SYMBOL == "MYL3" |
ttn2$SYMBOL == "ACTC1" 
,]
```

## Assess canonical transcript only

Assess canonical transcript only: the longest CDS (coding region of gene)       
MANE (via VEP) was assessed but only covered 16 of the 25 genes of interest.      
  
```{r filter canonical}
ttncan <- ttn2[which(ttn2$CANONICAL == "YES"),]
```

## Filter for rare variants (<0.1% (or missing) in UKB and gnomAD)     

gnomAD: The Genome Aggregation Database; 125,748 exomes from unrelated individuals sequenced as part of various disease-specific and population genetic studies, totalling 141,456 individuals. Some individuals with severe disease may still be included in the data sets, albeit likely at a frequency equivalent to or lower than that seen in the general population.     
  
```{r filter rare}
colnames(ttncan)[1] <- "ID" #was called #UploadedVariation by VEP
ttncan <- data.frame(ttncan)
ttncan2 <- join(ttncan, freq, type="left", by="ID")

#presence of "-" cause columns to not be numeric in R
ttncan2$gnomADg_faf95_nfe <- as.numeric(ttncan2$gnomADg_faf95_nfe)
ttncan2$gnomADg_faf95_afr <- as.numeric(ttncan2$gnomADg_faf95_afr)
ttncan2$gnomADg_faf95_amr <- as.numeric(ttncan2$gnomADg_faf95_amr)
ttncan2$gnomADg_faf95_eas <- as.numeric(ttncan2$gnomADg_faf95_eas)
ttncan2$gnomADg_faf95_sas <- as.numeric(ttncan2$gnomADg_faf95_sas)

#create a column for max faf95 from the five populations
ttncan2$gnomADg_faf95_popmax <- do.call('pmax',c(ttncan2[which( colnames(ttncan2)=="gnomADg_faf95_nfe" ):which( colnames(ttncan2)=="gnomADg_faf95_sas" )],list(na.rm=TRUE)))          
#if ran VEP as noted, nfe:sas in order

class(ttncan2$MAF)
#numeric
#MAF<0.1%
tmp <- ttncan2[which(ttncan2$MAF < 0.001),]

#gnomAD AF<0.1% or NA
tmp$gnomADg_AF <- as.numeric(tmp$gnomADg_AF)
ttncanrare <- tmp[which(tmp$gnomADg_AF < 0.001 | is.na(tmp$gnomADg_AF)),]
```

## Filter splice region variants    
### !Manual assessment required        

Identify those that are pathogenic/likely pathogenic for HCM in ClinVar and manually assess the ClinVar report for functional evidence of splice action;     
If splice acting evidence provided, renamed consequence as "splice_confirmed".    
If evidence is unclear, left consequence as is, e.g. "missense_variant,splice_region_variant".     
If evidence of alternative mechanism, renamed consequence without splice_region_variant flag to alternative mechanism only, e.g. "missense_variant".    

Assess the splice predictions of splice_region_variants that are found with another PAV flag,   
those that may be removed from analysis based on LOF flag but included if predicted to cause splicing:       
TNNT2 stop gained identified predicted to cause splicing, but is NMD competent so not included in analysis.  

Filter the splice_region_variants that are found with another non-PAV flag (e.g. intron, 3'UTR, 5'UTR, etc.):    
--> non-PAV-linked splice_region_variants filtered out that don't meet prediction thresholds of MaxEntScan_diff>3, ada_score>0.95, rf_score>0.75.   

```{r splice region}
#subset splice region variants
hmm <- ttncanrare[grepl("splice_region_variant", ttncanrare$Consequence),]

#are there any pathogenic for HCM with multiple submitters in ClinVar?
hmm2 <- hmm[which(hmm$ClinVar_CLNSIG == "Pathogenic" | hmm$ClinVar_CLNSIG == "Pathogenic/Likely_pathogenic" | hmm$ClinVar_CLNSIG == "Likely_pathogenic"),]

#HCM?
table(hmm2$ClinVar_CLNDN)
#all HCM

#Check ClinVar evidence strength for multiple submitters      
data.frame(hmm2$rf_score, hmm2$ada_score, hmm2$MaxEntScan_diff, hmm2$ClinVar, hmm2$SYMBOL, hmm2$Consequence, hmm2$ID)

#ClinVar_ID - Gene + variant type - Comment
#"42706" - MYBPC3 intron - confirmed splice disruption [shift in the reading frame, and premature termination of translation in exon 31 (Watkins et al. 1995)] + meets all thresholds - change to splice_confirmed
ttncanrare$Consequence[ttncanrare$ClinVar == "42706"] <- "splice_confirmed"
#"155808" - MYBPC3 intron - confirmed splice disruption [ splicing causing exon 29 skipping that results in a truncated protein, skips exon 29 (Rodriguez-Garcia et al., 2010; Crehalet et al., 2012)] + meets all thresholds - change to splice_confirmed
ttncanrare$Consequence[ttncanrare$ClinVar == "155808"] <- "splice_confirmed"
#"42556" - MYBPC3 intron - confirmed splice disruption [skips exon 17, leading to the introduction of a premature stop codon (Helms et al. 2014)] + meets all thresholds - change to splice_confirmed
ttncanrare$Consequence[ttncanrare$ClinVar == "42556"] <- "splice_confirmed"
#"8608" - MYBPC3 missense - confirmed splice disruption [skips exon 17,  and truncated protein product(Carrier_1997, Marston_2012, Helms_2014)] + meets all thresholds - change to splice_confirmed 
ttncanrare$Consequence[ttncanrare$ClinVar == "8608"] <- "splice_confirmed"
#"42792" - MYBPC3 missense - confirmed splice disruption [skips exon 6 , creates a truncated protein product (Sarikas et al., 2005; Helms et al., 2014)] + meets all thresholds - change to splice_confirmed
ttncanrare$Consequence[ttncanrare$ClinVar == "42792"] <- "splice_confirmed"
#"42784"  - MYBPC3 missense - has been shown in vitro (Crehalet et al., 2012) to skip exon 6 but another ClinVar report says the invitro assay is not representative + doesn't meet thresholds - leaving as is, will be included in analysis anyway
#IGNORE
#"378215" - MYH7 synonymous - confirmed splice disruption [inframe skipping of exon 38,  resulting in a transcript lacking exon 38 (Pajusalu et al., 2016; Fiorillo et al., 2016)] + meets all thresholds - change to splice_confirmed as indels are usually smaller than an entire exon.
ttncanrare$Consequence[ttncanrare$ClinVar == "378215"] <- "splice_confirmed"

#shorten all other splice_region_variants to those with computational evidence
#as.numeric (VEP calls NAs as "-")
ttncanrare$rf_score <- as.numeric(ttncanrare$rf_score)
ttncanrare$MaxEntScan_diff <- as.numeric(ttncanrare$MaxEntScan_diff)
ttncanrare$ada_score <- as.numeric(ttncanrare$ada_score)

#first filter for those meeting insilico for splicing
df <- ttncanrare %>%
  filter(Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,5_prime_UTR_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant,NMD_transcript_variant" | Consequence == "splice_region_variant,intron_variant,NMD_transcript_variant" | Consequence == "splice_region_variant,intron_variant,non_coding_transcript_variant" | Consequence == "splice_region_variant,non_coding_transcript_exon_variant") %>% 
  filter(MaxEntScan_diff<3 | ada_score<0.95 | rf_score<0.75) 

ttncanrare$found <- ttncanrare$ID %in% df$ID
ttncanrare2 <- ttncanrare[which(ttncanrare$found == "FALSE"),]
ttncanrare2$found <- NULL

#Then filter out those with NAs (can't be sure they meet insilico prediction thresholds)
df <- ttncanrare2 %>%
  filter(Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,5_prime_UTR_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant,NMD_transcript_variant" | Consequence == "splice_region_variant,intron_variant,NMD_transcript_variant" | Consequence == "splice_region_variant,intron_variant,non_coding_transcript_variant" | Consequence == "splice_region_variant,non_coding_transcript_exon_variant") %>%
  filter((is.na(rf_score)) | (is.na(MaxEntScan_diff)) | (is.na(ada_score)))

ttncanrare2$found <- ttncanrare2$ID %in% df$ID
ttncanrare2 <- ttncanrare2[which(ttncanrare2$found == "FALSE"),]
ttncanrare2$found <- NULL

write.table(ttncanrare2, "step1vars.txt", quote=F, row.names=F, sep="\t")
```

## Step 1: Find Genotype Negatives        

Identify carriers of rare protein altering variants in any of the 25 genes of interest: too inclusive but identifies controls/genotype negatives.  

```{r List Variants}
togo <- data.frame(ttncanrare2$ID)
write.table(togo, "IDsforextract.txt", quote=F, row.names=F)
```

```{bash Extract Variants}
plink --bfile ukb200exomehcm --extract IDsforextract.txt --recode A --output-chr M --out hcmvarsofinterest
```

```{r Find Carriers}
jabin <- fread("hcmvarsofinterest.raw", header=T)
jabin <- data.frame(jabin)

#Sample info always starts on 7th column in this data format
jabin$ref <- rowSums(jabin[7:ncol(jabin)] == "0", na.rm=T)
jabin$hets <- rowSums(jabin[7:ncol(jabin)-1] == "1", na.rm=T)
jabin$homs <- rowSums(jabin[7:ncol(jabin)-2] == "2", na.rm=T)
jabin$miss <- rowSums(is.na(jabin[7:ncol(jabin)-3]))

#list heterozygotes
hets <- jabin[which(jabin$hets > 0),]
hets_2 <- data.frame(hets$FID, hets$hets)
colnames(hets_2) <- c("eid", "count")
write.table(hets_2, "step1hets.txt", quote=F, row.names=F)

#list homozygotes
homs <- jabin[which(jabin$homs > 0),]
homs_2 <- data.frame(homs$FID)
colnames(homs_2) <- c("eid")
write.table(homs_2, "step1homs.txt", quote=F, row.names=F)

#assess variant missingness
ttncanrare2 <- fread("step1vars.txt", header=T, sep="\t")
ttncanrare2 <- data.frame(ttncanrare2)
rarevar <- join(ttncanrare2, miss, type="left", by="ID")
#which variants missing in greater than 10% of the population?
hmm <- rarevar[which(rarevar$N_MISS > (0.1*200643)),]
data.frame(hmm$ID, hmm$SYMBOL, hmm$Consequence, hmm$N_MISS)
#1 2:219418555:T:G        DES   missense_variant     199950
#2 7:128848680:G:C       FLNC   missense_variant      32467
#3 15:84817498:C:T      ALPK3   missense_variant      21543
#4 19:55156278:D:1      TNNI3 frameshift_variant      20996
#Leaving these four variants in to exclude individuals that carry these from being called as genotype negatives (even if sequencing quality is bad)
```

## Step 2:  
### !Manual assessment required        

Identify carriers of gene-specific consequences of the 8 sarcomeric genes:     
 
#### A - Potential "LoF" alleles:   
    stop_gained     
    splice_acceptor_variant
    splice_donor_variant  
    frameshift_variant     
    splice_confirmed        
    splice_region_variant (with additional insilico evidence)      
    start_lost (with addition sequence evidence of LoF)     
    ==> use group (i) as LoF, (ii) as PAV.    
    ==> remove group (iii).    
    ==> assess group (iv).    
    
#### B - Altered protein:   
    missense_variant     
    inframe_insertion   
    inframe_deletion   
    stop_lost  (with addition sequence evidence)      
    start_lost  (with addition sequence evidence)     
    Any of the LoF variant classes if predicted to escape NMD by LOFTEE, flag as "NMDi".      
    ==> remove group (i).        
    ==> use group (ii).           
    ==> assess group (iii) and (iv).    

##### 4 bins incorporating LOFTEE:    
(i) introduces a PTC, and predicted to lead to NMD (= HC LOFTEE but not necessarily NMD incompetent).    
(ii) introduces a PTC, but predicted to escape NMD = "NMDi" (= positional annotation; 55bp into penultimate exon - NMD incompetent).   
(iii) those likely mis-annotated, or not in a relevant exon, so no PTC introduced.   
(iv) those flagged as "low confidence (LC)" with other LOFTEE flags e.g. NAGNAG.      

#### Final filter based on consequence terms identified
A_List=c("frameshift_variant", "frameshift_variant,start_lost", "splice_donor_variant", "splice_acceptor_variant", "splice_confirmed", "splice_donor_variant,coding_sequence_variant", "splice_region_variant,intron_variant", "splice_region_variant,synonymous_variant",  "stop_gained", "stop_gained,frameshift_variant","stop_gained,splice_region_variant","frameshift_variant_NMDi","splice_acceptor_variant_NMDi","splice_donor_variant_NMDi","splice_donor_variant_NMDi,coding_sequence_variant","splice_donor_variant,intron_variant","splice_region_variant_NMDi,intron_variant","splice_region_variant_NMDi,synonymous_variant","stop_gained_NMDi", "splice_donor_variant,intron_variant")       
        
B_List=c("missense_variant", "missense_variant,splice_region_variant", "inframe_deletion", "inframe_insertion", "start_lost","start_lost,splice_region_variant", "stop_lost", "splice_acceptor_variant_NMDi", "splice_donor_variant_NMDi", "splice_donor_variant_NMDi,coding_sequence_variant", "splice_region_variant_NMDi,intron_variant", "frameshift_variant_NMDi", "stop_gained_NMDi", "splice_region_variant_NMDi,synonymous_variant")        

MYH7 | B_List.   
MYBPC3 | A_List + B_List.  
TNNT2 | B_List.     
TNNI3 | B_List.   
TPM1 | B_List.    
MYL2 | B_List.  
MYL3 | B_List.  
ACTC1 | B_List.   

Plus meets a gnomAD filter allele frequency (FAF) threshold for HCM (0.00004).
  To identify variants above the maximum credible allele frequency for HCM, FAF is the highest filtering allele frequency across all gnomAD populations (“popmax”) for each variant.  

```{r ant8 - Step 2}
#read in variants from step before
ttncanrare2 <- fread("step1vars.txt")
ttncanrare2 <- data.frame(ttncanrare2)

#add bim file to get more variant info like position (POS)
doneant <- join(ttncanrare2, bim, type="left", by="ID")

#keep 8 sarcomeric genes only
doneant <- doneant[
doneant$SYMBOL == "MYH7" |
doneant$SYMBOL == "MYBPC3"|
doneant$SYMBOL == "TNNT2"  |
doneant$SYMBOL == "TNNI3" |
doneant$SYMBOL == "TPM1" |
doneant$SYMBOL == "MYL2" |
doneant$SYMBOL == "MYL3" |
doneant$SYMBOL == "ACTC1" ,
]

#What variants does LOFTEE pick up, are they NMD incompetent?
table(doneant$LoF_filter, doneant$SYMBOL)
table(doneant$LoF_flags, doneant$SYMBOL)
#1 MYL3 3'UTR SPLICE
#1 MYBPC3 NAGNAG SITE
doneant$Consequence[doneant$LoF_flags == "NAGNAG_SITE"] <- "inframe_insertion"
doneant$Consequence[doneant$LoF_filter == "3UTR_SPLICE"] <- "splice_donor_variant_LC"
#1 NAGNAG site = splice acceptor site rescued by inframe acceptor site (low confidence)
#3'UTR SPLICE = splice donor with pLof low confidence

#What LOF variants does LOFTEE miss that could be positionally NMD incompetent?
hmm <- doneant %>%
  filter(SYMBOL == "MYH7") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#MYH7 <	23,413,759 + 55
hmm2 <- hmm[which(hmm$POS < 23413814),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "14:23412872:C:A" & doneant$SYMBOL == "MYH7"] <- "splice_acceptor_variant_NMDi"
doneant$Consequence[doneant$ID == "14:23413757:A:G" & doneant$SYMBOL == "MYH7"] <- "splice_donor_variant_NMDi"
doneant$Consequence[doneant$ID == "14:23413759:C:T" & doneant$SYMBOL == "MYH7"] <- "splice_region_variant_NMDi,synonymous_variant"

hmm <- doneant %>%
  filter(SYMBOL == "MYBPC3") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#MYBPC3 <	47,332,072 + 55
hmm2 <- hmm[which(hmm$POS < 47332127),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "11:47332075:G:A" & doneant$SYMBOL == "MYBPC3"] <- "stop_gained_NMDi"

hmm <- doneant %>%
  filter(SYMBOL == "TNNT2") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#TNNT2 < 201,359,623 + 55
hmm2 <- hmm[which(hmm$POS < 201359678),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "1:201359258:G:T" & doneant$SYMBOL == "TNNT2"] <- "splice_region_variant_NMDi,intron_variant"
doneant$Consequence[doneant$ID == "1:201359217:C:T" & doneant$SYMBOL == "TNNT2"] <- "stop_gained_NMDi"
doneant$Consequence[doneant$ID == "1:201359651:G:A" & doneant$SYMBOL == "TNNT2"] <- "stop_gained_NMDi"
doneant$Consequence[doneant$ID == "1:201359665:T:G" & doneant$SYMBOL == "TNNT2"] <- "splice_acceptor_variant_NMDi"

hmm <- doneant %>%
  filter(SYMBOL == "TNNI3") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#TNNI3 < 	55,154,030 + 55
hmm2 <- hmm[which(hmm$POS < 55154085),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "19:55154026:T:A" & doneant$SYMBOL == "TNNI3"] <- "splice_region_variant_NMDi,intron_variant"

#Ignore TPM1 - one exon only

hmm <- doneant %>%
  filter(SYMBOL == "MYL2") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#MYL2 < 110,913,096 + 55
hmm2 <- hmm[which(hmm$POS < 110913151),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "12:110911146:D:1" & doneant$SYMBOL == "MYL2"] <- "frameshift_variant_NMDi"
doneant$Consequence[doneant$ID == "12:110911176:C:G" & doneant$SYMBOL == "MYL2"] <- "splice_acceptor_variant_NMDi"
doneant$Consequence[doneant$ID == "12:110911176:C:T" & doneant$SYMBOL == "MYL2"] <- "splice_acceptor_variant_NMDi"
doneant$Consequence[doneant$ID == "12:110913094:D:3" & doneant$SYMBOL == "MYL2"] <- "splice_donor_variant_NMDi,coding_sequence_variant"
doneant$Consequence[doneant$ID == "12:110913095:C:A" & doneant$SYMBOL == "MYL2"] <- "splice_donor_variant_NMDi"

hmm <- doneant %>%
  filter(SYMBOL == "MYL3") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#MYL3 <	46,858,244 + 55
hmm2 <- hmm[which(hmm$POS < 46858299),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "3:46858231:C:T" & doneant$SYMBOL == "MYL3"] <- "splice_region_variant_NMDi,3_prime_UTR_variant"

hmm <- doneant %>%
  filter(SYMBOL == "ACTC1") %>%
  filter(Consequence == "frameshift_variant" | Consequence == "frameshift_variant,start_lost" | Consequence == "splice_acceptor_variant" | Consequence == "splice_confirmed" | Consequence == "splice_donor_variant" | Consequence == "splice_donor_variant,coding_sequence_variant" | Consequence == "splice_donor_variant,intron_variant" | Consequence == "splice_region_variant,3_prime_UTR_variant" | Consequence == "splice_region_variant,intron_variant" | Consequence == "splice_region_variant,synonymous_variant" | Consequence == "start_lost" | Consequence == "start_lost,splice_region_variant" | Consequence == "stop_gained" | Consequence == "stop_gained,frameshift_variant" | Consequence == "stop_gained,splice_region_variant")

#ACTC1 < 34,791,114 + 55
hmm2 <- hmm[which(hmm$POS < 34791169),]
data.frame(hmm2$ID, hmm2$Consequence, hmm2$POS)

doneant$Consequence[doneant$ID == "15:34790493:D:2" & doneant$SYMBOL == "ACTC1"] <- "frameshift_variant_NMDi"

#Assess presence of protein altering variants?
hmm <- doneant %>%
  filter(Consequence == "protein_altering_variant")
#none found at this stage

#Assess presence of coding_sequence_variant?
hmm <- doneant %>%
  filter(Consequence == "coding_sequence_variant")
#none found at this stage

#Manually assess the ensembl sequence for start and stop codons in each gene for another inframe start/stop nearby (before or after variant)
#In summary for the variants seen:
#Start codons cause indels for MYBPC3 (x3), MYH7, TPM1 (x2), MYL2 (x2), TNNT2, and truncations for ACTC1 and MYL3 (these two will be excluded from our HCM pipeline).
#Stop codons cause inframe insertion for MYH7 and TNNI3.

A_List=c("frameshift_variant", "frameshift_variant,start_lost", "splice_donor_variant", "splice_acceptor_variant", "splice_confirmed", "splice_donor_variant,coding_sequence_variant", "splice_region_variant,intron_variant", "splice_region_variant,synonymous_variant",  "stop_gained", "stop_gained,frameshift_variant", "stop_gained,splice_region_variant","frameshift_variant_NMDi","splice_acceptor_variant_NMDi","splice_donor_variant_NMDi","splice_donor_variant_NMDi,coding_sequence_variant","splice_donor_variant,intron_variant","splice_region_variant_NMDi,intron_variant","splice_region_variant_NMDi,synonymous_variant","stop_gained_NMDi", "splice_donor_variant,intron_variant","missense_variant,splice_region_variant")

B_List=c("missense_variant", "missense_variant,splice_region_variant", "inframe_deletion", "inframe_insertion", "start_lost", "start_lost,splice_region_variant", "stop_lost", "splice_acceptor_variant_NMDi", "splice_donor_variant_NMDi", "splice_donor_variant_NMDi,coding_sequence_variant", "splice_region_variant_NMDi,intron_variant", "frameshift_variant_NMDi", "stop_gained_NMDi", "splice_region_variant_NMDi,synonymous_variant")

doneant8 <- doneant[
doneant$SYMBOL == "MYH7" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "MYBPC3" & grepl(paste(A_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "MYBPC3" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "TNNT2" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "TNNI3" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "TPM1" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "MYL2" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "MYL3" & grepl(paste(B_List, collapse="|"), doneant$Consequence) |
doneant$SYMBOL == "ACTC1" & grepl(paste(B_List, collapse="|"), doneant$Consequence),
]  

dim(doneant8)
#1786 variants

### !Manual Assessment
#Ensure removal of:
#splice_region_variant_NMDi,3_prime_UTR_variant
#start_lost in ACTC1 and MYL3
#splice_donor_variant_LC 

class(doneant8$gnomADg_faf95_popmax)
#numeric
done3 <- doneant8[which(doneant8$gnomADg_faf95_popmax < 0.00004 | is.na(doneant8$gnomADg_faf95_popmax)),]
write.table(done3, "step2vars.txt", quote=F, row.names=F, sep="\t")

togo <- data.frame(done3$ID)
write.table(togo, "IDsforextract_2.txt", quote=F, row.names=F, col.names=F)
```

```{bash}
plink --bfile ukb200exomehcm --extract IDsforextract_2.txt --recode A --real-ref-alleles --output-chr M --out step2varsofinterest
```

```{r}
jabin <- fread("step2varsofinterest.raw", header=T)
jabin <- data.frame(jabin)

jabin$ref <- rowSums(jabin[7:ncol(jabin)] == "0", na.rm=T)
jabin$hets <- rowSums(jabin[7:ncol(jabin)-1] == "1", na.rm=T)
jabin$homs <- rowSums(jabin[7:ncol(jabin)-2] == "2", na.rm=T)
jabin$miss <- rowSums(is.na(jabin[7:ncol(jabin)-3]))

#list hets
hets <- jabin[which(jabin$hets > 0),]
hets_2 <- data.frame(hets$FID, hets$hets)
colnames(hets_2) <- c("eid", "count")
write.table(hets_2, "step2hets.txt", quote=F, row.names=F)

#count homs
table(jabin$homs)
#    0
#200643
#none by this point
```

## Step 3A:
### !Manual Assessment required

as (2B) above plus meets criteria would report as P/LP if you saw it in HCM - annotated pathogenic/likely pathogenic using Cardioclassifier (CC - https://www.cardioclassifier.org/) in a autosomal dominant fashion for HCM (e.g. assessed as heterozygotes) and additional manual curation from literature (Walsh et al. 2019; Table S6).     
  CC supports disease specific interpretation of genetic variants in genes associated with Inherited Cardiac Conditions (ICCs), according to guidelines released by the American College of Medical Genetics and Genomics (2015).
  The variants included in this step from Walsh et al. 2019 were called as pathogenic/likely-pathogenic and had a flag of ACMG rules PP5, PP4, PP1, PM6, PM3, PS3, PS2, PM6, PS2, PP1.  
  
GrCh38 variants were lifted over to GrCh37 for CardioClassifier.       

```{r step 3A}
done3 <- fread("step2vars.txt", header=T, sep="\t")
done3 <- data.frame(done3)

forcc <- data.frame(done3$ID, done3$CHROM, done3$POS, done3$REF, done3$ALT)
forcc$forlift <- paste0("chr",forcc$done3.CHROM, ":", forcc$done3.POS,"-", forcc$done3.POS)
write.table(forcc, "hcm_forcc_grch38.txt", quote=F, row.names=F, col.names=F)

#lift it over to grch37 on UCSC (https://genome.ucsc.edu/cgi-bin/hgLiftOver) or liftover script = HCM_CC_input.txt.    
#CardioClassifier output = HCM_CC_out_het.txt.    

#join grch37 file with cardioclassifier input file and output file
#then join this dataset with the list of 2B variants
lift <- read.table("HCM_CC_input.txt", header=T)
colnames(lift) <- c("Chr","Pos", "Ref", "Alt")
forcc <- read.table("hcm_forcc_grch38.txt", header=F)
colnames(forcc) <- c("ID","Chr","POS", "REF", "ALT","forlift")
tgthr <- data.frame(forcc,lift)
cc <- fread("HCM_CC_out_het.txt")
done <- join(tgthr, cc, type="left", by=c("Chr", "Pos", "Ref", "Alt"))
colnames(done)[2] <- "CHROM"
colnames(done)[8] <- "grch37"
done4 <- join(done3, done, type="left", by=c("CHROM", "POS", "REF", "ALT"))

#Filter for those that are P/LP in CC
done5 <- done4[which(done4$ACMG_class == "Likely Pathogenic" | done4$ACMG_class == "Pathogenic"),]

#incorporate cosegregation data and manual variant curation from Walsh et al. 2019
coseg <- read.table("coseg.txt", header=T)
colnames(coseg)[1] <- "SYMBOL"
colnames(coseg)[2] <- "HGVSc_2"
#change frameshifts to c.X_Xdel only
coseg$HGVSc_2 <- as.character(coseg$HGVSc_2)
coseg$HGVSc_2[coseg$HGVSc_2 == "c.1357_1358delCC"] <- "c.1357_1358del"
coseg$HGVSc_2[coseg$HGVSc_2 == "c.2864_2865delCT"] <- "c.2864_2865del"

#reshape the VEP output of HGVSc
library(splitstackshape)
done5B2 <- cSplit(done3, 'HGVSc', sep=":", type.convert=FALSE)

#join coseg data with large dataframe and see if IDs in step 2B variants
done5B3 <- join(coseg, done5B2, type="left", by=c("HGVSc_2"))
togo <- data.frame(done5B3[1:5]) 
done3$found <- done3$ID %in% togo$ID  
right <- done3[which(done3$found == "TRUE"),]
right$found <- NULL

#join the cardioclassifier variant list with the 2B variant list and remove duplicates
done5Ba <- rbind.fill(done5, right)
done5 <- done5Ba[!duplicated(done5Ba$ID),]

write.table(done5, "step3Avars.txt", quote=F, row.names=F, sep="\t")

togo <- data.frame(done5$ID)
write.table(togo, "IDsforextract_3A.txt", quote=F, row.names=F, col.names=F)
```

```{bash}
plink --bfile ukb200exomehcm --extract IDsforextract_3A.txt --recode A --real-ref-alleles --output-chr M --out step3avarsofinterest
```

```{r}
jabin <- fread("step3avarsofinterest.raw", header=T)
jabin <- data.frame(jabin)

jabin$ref <- rowSums(jabin[7:ncol(jabin)] == "0", na.rm=T)
jabin$hets <- rowSums(jabin[7:ncol(jabin)-1] == "1", na.rm=T)
jabin$homs <- rowSums(jabin[7:ncol(jabin)-2] == "2", na.rm=T)
jabin$miss <- rowSums(is.na(jabin[7:ncol(jabin)-3]))

#list hets
hets <- jabin[which(jabin$hets > 0),]
hets_2 <- data.frame(hets$FID, hets$hets)
colnames(hets_2) <- c("eid", "count")
write.table(hets_2, "step3Ahets.txt", quote=F, row.names=F)
```

## Step 3B:

as (2B*) above plus has been found in someone with HCM and deemed pathogenic/likely pathogenic based on the evidence provided to ClinVar.     
  ClinVar processes submissions reporting variants found in patient samples, assertions made regarding their clinical significance, information about the submitter, and other supporting data.    
  
```{r step 3B}
#Which are found in HCM in ClinVar?
done3 <- fread("step2vars.txt", header=T, sep="\t")
done3 <- data.frame(done3)

hyp <- c("Hypertrophic", "hypertrophic")
done5Ba <- done3[grepl(paste(hyp, collapse="|"), done3$ClinVar_CLNDN),]

done5Bb <- done5Ba[which(done5Ba$ClinVar_CLNSIG == "Pathogenic" | done5Ba$ClinVar_CLNSIG == "Likely_pathogenic" | done5Ba$ClinVar_CLNSIG == "Pathogenic/Likely_pathogenic"),]

done5B <- done5Bb[which(done5Bb$ClinVar_CLNREVSTAT == "criteria_provided,_multiple_submitters,_no_conflicts" | done5Bb$ClinVar_CLNREVSTAT == "reviewed_by_expert_panel" | done5Bb$ClinVar_CLNREVSTAT == "practice_guideline"),]

write.table(done5B, "step3Bvars.txt", quote=F, row.names=F, sep="\t")

togo <- data.frame(done5B$ID)
write.table(togo, "IDsforextract_3B.txt", quote=F, row.names=F, col.names=F)
```

```{bash}
plink --bfile ukb200exomehcm --extract IDsforextract_3B.txt --recode A --real-ref-alleles --output-chr M --out step3bvarsofinterest
```

```{r}
jabin <- fread("step3bvarsofinterest.raw", header=T)
jabin <- data.frame(jabin)

jabin$ref <- rowSums(jabin[7:ncol(jabin)] == "0", na.rm=T)
jabin$hets <- rowSums(jabin[7:ncol(jabin)-1] == "1", na.rm=T)
jabin$homs <- rowSums(jabin[7:ncol(jabin)-2] == "2", na.rm=T)
jabin$miss <- rowSums(is.na(jabin[7:ncol(jabin)-3]))

#list hets
hets <- jabin[which(jabin$hets > 0),]
hets_2 <- data.frame(hets$FID, hets$hets)
colnames(hets_2) <- c("eid", "count")
write.table(hets_2, "step3Bhets.txt", quote=F, row.names=F)
```

## Summary Table: summarytable_meta.txt      

Took a curated "metadata" UKB phenotype file that contains summary information on UKB participants. 
Subsetted the data for those with Exome data.
Added 7 columns explaining the steps above (data is binary; 0 - no variant, 1 - has variant):    
Step 0 = 1:  people have no rare PAV = 1  
Step 1 = 1:  people any rare PAV (syndromic or not)  
Step 2 = 1:  people have a rare PAV (FAF for HCM) in the 8 sarcomere genes 
Step 3A = 1:  people have a P/LP (CC) rare (FAF) variant in the 8 sarcomere genes  
Step 3B = 1:  people have a ClinVar HCM rare (FAF) P/LP variant in the 8 sarcomere genes  
  PAV = protein altering variant  
  FAF = filtering allele frequency  
  P/LP = flagged as pathogenic or likely pathogenic  
  CC = CardioClassifier  
The columns summarise all homozygotes and heterozygotes into one.   

```{r summarise table}
meta$hasExomeData200k <- meta$eid %in% fam$V1
cmrnex <- meta[which(meta$hasExomeData200k == "TRUE"),]

#READ IN
hom_step_1 <- read.table("step1homs.txt", header=T)
het_step_1 <- read.table("step1hets.txt", header=T)
het_step_2 <- read.table("step2hets.txt", header=T)
het_step_3A <- read.table("step3Ahets.txt", header=T)
het_step_3B <- read.table("step3Bhets.txt", header=T)

#Check colnames
names(het_step_1)
names(het_step_2)
names(het_step_3A)
names(het_step_3B)
names(hom_step_1)

#STEP 1
tmp1 <- cmrnex$eid %in% hom_step_1$eid
tmp2 <- cmrnex$eid %in% het_step_1$eid
tmp <- data.frame(tmp1, tmp2)
tmp$HCMstep1 <- do.call('pmax',c(tmp[,1:2],list(na.rm=TRUE)))

#STEP 2
tmp$HCMstep2 <- as.numeric(cmrnex$eid %in% het_step_2$eid)

#STEP 3A
tmp$HCMstep3A <- as.numeric(cmrnex$eid %in% het_step_3A$eid)

#STEP 3B
tmp$HCMstep3B <- as.numeric(cmrnex$eid %in% het_step_3B$eid)

#STEP 0
names(tmp)
tmp$HCMstep0 <- do.call('pmax',c(tmp[,c(3,4,5,6)],list(na.rm=TRUE)))
tmp$HCMstep0 <- gsub("1", 5, tmp$HCMstep0)
tmp$HCMstep0 <- gsub("0", 1, tmp$HCMstep0)
tmp$HCMstep0 <- gsub("5", 0, tmp$HCMstep0)
#people with 1 in flag_step0 don't have a PAV in any of the HCM genes of interest. 

tmp3 <- tmp[,c(7,3,4,5,6)]
final <- data.frame(cmrnex, tmp3)
dim(final)
#200632
head(final)
write.table(final, "summarytable_meta.txt", quote=F, row.names=F, sep="\t")
```

END.
